### 2022-03-10 omicron-generation.tex
more edits

We adjust for observed growth rate of each variant by taking the backward incubation-period distributions estimated by \cite{backer2021omicron} and calculating the corresponding forward incubation-period distributions by inverting \eref{backward}:
\begin{equation}
f_I(\tau) = \frac{\exp(r\tau) b_I(\tau) }{\int_0^\infty \exp(rx) b_I(x)\dx}.
\label{eq:forward}
\end{equation}
We model the backward incubation-period distribution $b_I(\tau)$ using a Weibull distribution based on the assumptions of \cite{backer2021omicron}.
To account for uncertainties in the original parameter estimates, we rely on a sampling scheme, similar to the one we used for the growth rate analysis (in Section 2.2).
First, we approximate the previously inferred posterior distributions of the shape and scale parameters of the Weibull distribution using a lognormal distribution---we parameterize the lognormal distribution such that (i) its median matches the median estimates and (ii) the probability that a random variable following the specified Gamma distribution falls between the lower and upper credible limits matches 95\% \citep{park2020reconciling}.
\jd{Why do we match a mean to a median?, sounds crazy.} \swp{I changed to lognormal and now match median. I was just doing something quick and dirty when I did this...}
We draw 1000 samples of the shape and scale parameters (for the backward distribution $b_I(\tau)$) from the specified lognormal distributions and estimate the corresponding forward distribution using \eref{forward}.
We take 95\% equi-tailed quantiles to obtain 95\% confidence intervals.

@@

Dynamical biases in the serial-interval distributions are more complex because the serial interval depends on the incubation periods of the infector and the infectee as well as the generation interval between them.
For example, \cite{backer2021omicron} measured the forward serial-interval distributions from cohorts of infectors who developed symptoms during the same week.
In this case, the forward serial interval $\tau_s$ can be expressed in the form \citep{park2021forward}:
\begin{equation}
\tau_s =  - \tau_{i, 1} + \tau_{g} + \tau_{i, 2},
\label{eq:serial}
\end{equation}
where $\tau_{i, 1}$ represents the backward incubation period of the infector (because all infectors developed symptoms at the same time), $\tau_{g|i,1}$ \jd{why does notation here not match equation?} represents the forward generation interval, and $\tau_{i, 2},$ represents the forward incubation period of the infectee.
When incidence of infection is growing exponentially, infectors in a given cohort are more likely to have shorter incubation periods than their infectees on average, $\mathbb E[\tau_{i, 1}] < \mathbb E[\tau_{i, 2}]$, causing the mean forward serial interval to be longer than the mean forward generation interval ($\mathbb E[\tau_s] > \mathbb E[\tau_{g}]$).

We further note that the mean forward generation interval $\mathbb E[\tau_{g}]$ defined in \eref{serial} is also likely shorter than the mean forward generation interval $\mathbb E[\hat{\tau}_{g}]$ that we wish to estimate because infectors with shorter incubation periods (due to exponential growth) also transmit faster.
To illustrate this point, we write $f_{G|I}(\tau_g|\tau_{i,1})$ to represent forward generation-interval distribution conditional on the incubation period.
Then, we can see that the mean forward generation interval $\mathbb E[\tau_{g}]$ defined in \eref{serial} depends on the exponential growth rate:

\begin{align}
\mathbb E[\tau_{g}] &= \int_0^\infty \int_0^\infty x_g f_{G|I}(x_g|x_{i,1}) b_I(x_{i,1}) \dd x_g x_{i, 1}\\
&= \frac{\int_0^\infty \int_0^\infty x_g \exp(-rx_{i,1})  f_{G|I}(x_g|x_{i,1}) f_I(x_{i,1}) \dd x_g x_{i, 1}}{\int_0^\infty  \exp(-r x) f_I(x) \dx}
\end{align}

\jd{What's going on here? I'm having trouble wading through the equations, but I guess the idea is that $\tau_g$ is the mean generation interval for a symptom-based cohort, and that what we want is the mean for an infection-based cohort. Can we say that more clearly? Can we also work on accompanying notation? In particular, I don't like giving the weird one the more-normal name.}
On the other hand, the mean forward generation interval of our interest \jd{“of our interest” requires some explanation, and maybe some thought} does not depend on the exponential growth rate:
\begin{equation}
\mathbb E[\hat{\tau}_{g}] = \int_0^\infty \int_0^\infty x_g f_{G|I}(x_g|x_{i,1}) f_I(x_{i,1}) \dd x_g x_{i, 1}.\\
\label{eq:meangen}
\end{equation}
Therefore, while we generally expect the mean forward generation interval to be shorter than the mean forward serial interval for a growing epidemic, correlation between symptom onset and transmissibility can also have opposite effects.
\jd{I'm not quite following here. Can we disentangle the two effects a bit better?}

\jd{Didn't we just finish modeling incubation periods with both Weibull and lognormal? Is there a simpler way?}
Here, we model forward incubation-period $f_I(\tau)$ and generation-interval $f_G(\tau)$ distributions with lognormal distributions.
Both distributions are parameterized by log means, $\mu_I$ and $\mu_G$, and log variances, $\sigma_I^2$ and $\sigma_G^2$.
Then, given the correlation coefficient $\rho$, the forward generation-interval distribution conditional conditional on the incubation period $f_{G|I}(\tau|\tau_{i,1})$ has a mean of $\mu_G + \sigma_G \rho (\log(\tau_{i, 1}) - \mu_I)/\sigma_I$ and a log variance of $\sigma_G^2 (1 - \rho^2)$.
Assuming that the incidence of infection is changing exponentially at rate $r$, the forward serial-interval distribution $f_S(\tau)$ for a cohort of infectors who developed symptoms at time $t = 0$ can be calculated by integrating across infection time of the infector $\alpha_1 < 0$ and of the infectee $\alpha_2 > \alpha_1$ \citep{park2021forward}:
\begin{equation}
f_S(\tau) = \frac{1}{\phi} \int_{-\infty}^0\int_{\alpha_1}^\tau \exp(r \alpha_1) f_{G|I}(\alpha_2 - \alpha_1|- \alpha_1) f_I(- \alpha_1) f_I(\tau - \alpha_2) \dd \alpha_2 \dd \alpha_1,
\label{eq:forwardserial}
\end{equation}
where $\phi$ represents the normalization constant to satisfy $\int f_S(x) \dx = 1$;
$-\alpha_1$ corresponds to the incubation period of the infector;
$\alpha_2 - \alpha_1$ corresponds to the generation interval;
and $\tau - \alpha_2$ corresponds to the incubation period of the infectee.

>>>>

We adjust for observed growth rate of each variant by taking the backward incubation-period distributions estimated by \cite{backer2021omicron} and calculating the corresponding forward incubation-period distributions by inverting \eref{backward}:
\begin{equation}
f_I(\tau) = \frac{\exp(r\tau) b_I(\tau) }{\int_0^\infty \exp(rx) b_I(x)\dx}.
\label{eq:forward}
\end{equation}
We model the backward incubation-period distribution $b_I(\tau)$ using a Weibull distribution based on the assumptions of \cite{backer2021omicron}.
To account for uncertainties in the original parameter estimates, we rely on a sampling scheme, similar to the one we used for the growth rate analysis (in Section 2.2).
First, we approximate the previously inferred posterior distributions of the shape and scale parameters of the Weibull distribution using a lognormal distribution---we parameterize the lognormal distribution such that (i) its median matches the median of the posterior distributions and (ii) the probability that a random variable following the specified Gamma distribution falls between the lower and upper credible limits matches 95\% \citep{park2020reconciling}.
We draw 1000 samples of the shape and scale parameters (for the backward distribution $b_I(\tau)$) from the specified lognormal distributions and estimate the corresponding forward distribution using \eref{forward}.
We take 95\% equi-tailed quantiles to obtain 95\% confidence intervals.

@@

Dynamical biases in the serial-interval distributions are more complex because the serial interval depends on the incubation periods of the infector and the infectee as well as the generation interval between them.
For example, \cite{backer2021omicron} measured the forward serial-interval distributions from cohorts of infectors who developed symptoms during the same week.
In this case, the forward serial interval $\tau_s$ can be expressed in the form \citep{park2021forward}:
\begin{equation}
\tau_s =  - \tau_{i, 1} + \tau_{g, \mathrm{symp}} + \tau_{i, 2},
\label{eq:serial}
\end{equation}
where $\tau_{i, 1}$ represents the backward incubation period of the infector (because all infectors developed symptoms at the same time), and $\tau_{i, 2},$ represents the forward incubation period of the infectee.
Here, $\tau_{g, \mathrm{symp}}$ represents the generation interval between the infector and the infectee; we use the subscript $\mathrm{symp}$ to indicate that these generation intervals are measured from infectors who developed symptoms at the same time.

The symptom-based generation-interval distribution describing \eref{serial} gives a biased picture because infectors who developed symptoms at the same time will have shorter incubation periods (when epidemic is growing) and therefore transmit earlier.
This symptom-based generation-interval distribution depends on the backward incubation-period distribution:
\begin{align}
f_{G, \mathrm{symptom}}(\tau) &= \int_0^\infty f_{G|I}(\tau|x) b_I(x) \dx,
\end{align}
where $f_{G|I}(\tau|x)$ represents the forward generation-interval distribution conditional on a known value of the incubation period, $x$.
Instead, the forward generation-interval distribution measured from a cohort of individuals who were infected at the time is expected to provide reliable estimates of the distribution across individuals (because their incubation-period distribution is expected to remain constant over time):
\begin{align}
f_{G, \mathrm{inf}}(\tau) &= \int_0^\infty f_{G|I}(\tau|x) f_I(x) \dx.
\end{align}

When epidemic is growing is exponentially, we have two opposing effects affecting the relationship between the mean serial and generation interval.
First, infectors in a given cohort are more likely to have shorter incubation periods than their infectees on average, $\mathbb E[\tau_{i, 1}] < \mathbb E[\tau_{i, 2}]$, causing the mean forward serial interval to be longer than the mean symptom-based generation interval ($\mathbb E[\tau_s] > \mathbb E[\tau_{g, \mathrm{symp}}]$).
Second, the mean symptom-based generation interval will be shorter than the mean infection-based generation interval: $E[\tau_{g, \mathrm{inf}} > \mathbb E[\tau_{g, \mathrm{symp}}]$.
Therefore, the difference between the mean serial interval and the mean infection-based generation interval is difficult to predict in general; in most cases, however, we expect the former effect to dominate, causing the mean serial interval to be longer than the mean infection-based generation interval: $\mathbb E[\tau_s] > E[\tau_{g, \mathrm{inf}}$ \citep{park2021forward}.
Hereafter, we use the term ``forward generation-interval'' to refer to the infection-based generation-interval distribution (measured from a cohort of infectors who were infected at the same infection time) for simplicity and drop the subscript $\mathrm{inf}$.

Here, we model forward incubation-period $f_I(\tau)$ and generation-interval $f_G(\tau)$ distributions with lognormal distributions.
Both distributions are parameterized by log means, $\mu_I$ and $\mu_G$, and log variances, $\sigma_I^2$ and $\sigma_G^2$.
Then, given the correlation coefficient $\rho$, the forward generation-interval distribution conditional conditional on the incubation period $f_{G|I}(\tau|\tau_{i,1})$ has a mean of $\mu_G + \sigma_G \rho (\log(\tau_{i, 1}) - \mu_I)/\sigma_I$ and a log variance of $\sigma_G^2 (1 - \rho^2)$.
Assuming that the incidence of infection is changing exponentially at rate $r$, the forward serial-interval distribution $f_S(\tau)$ for a cohort of infectors who developed symptoms at time $t = 0$ can be calculated by integrating across infection time of the infector $\alpha_1 < 0$ and of the infectee $\alpha_2 > \alpha_1$ \citep{park2021forward}:
\begin{equation}
f_S(\tau) = \frac{1}{\phi} \int_{-\infty}^0\int_{\alpha_1}^\tau \exp(r \alpha_1) f_{G|I}(\alpha_2 - \alpha_1|- \alpha_1) f_I(- \alpha_1) f_I(\tau - \alpha_2) \dd \alpha_2 \dd \alpha_1,
\label{eq:forwardserial}
\end{equation}
where $\phi$ represents the normalization constant to satisfy $\int f_S(x) \dx = 1$;
$-\alpha_1$ corresponds to the incubation period of the infector;
$\alpha_2 - \alpha_1$ corresponds to the generation interval;
and $\tau - \alpha_2$ corresponds to the incubation period of the infectee.
